---
title: "Get started"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Get started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r include=FALSE}
knitr::opts_chunk$set(collapse  = TRUE,
                      comment   = "#>",
                      out.width = "100%",
                      dpi       = 96,
                      fig.align = "center")
```


The aim of the package `bridge` is to provide tools to work with directed (asymetric) 
spatial networks. It allows users to easily create and visualize network objects, 
like **nodes list**, **edges lists**, **connectivity matrix**, 
**nodes by edges matrix**, and **edges weights vector**. These objects can be used later in 
Asymetric Eigenvector Maps (AEM, Blanchet _et al._ 2008) available in the package
[`adespatial`](https://cran.r-project.org/package=adespatial) (Dray _et al._ 2022).

This package has been developed for the 
[FRB-CESAB](https://www.fondationbiodiversite.fr/en/about-the-foundation/le-cesab/) 
working group 
[Bridge](https://www.fondationbiodiversite.fr/en/the-frb-in-action/programs-and-projects/le-cesab/bridge/) 
that aims to better understand the role of local and regional environmental 
factors in shaping the taxonomic and functional diversity of plant communities 
established along river corridors, roadside corridors and cultivated field 
margins.


\


The package `bridge` also provides some tools to work with Moran's Eigenvector Maps 
(MEM, Dray _et al._ 2006) but this vignette only focuses on 
**directed spatial network** and AEM.


\



```{r 'setup'}
library("bridge")
library("ggplot2")
```


\


First let's define a custom theme for the `ggplot2` visualizations.


```{r 'ggplot-theme'}
custom_theme <- function() {
  theme_light() + 
  theme(plot.title   = element_text(face = "bold", family = "serif", size = 18),
        plot.caption = element_text(face = "italic", family = "serif"),
        axis.title   = element_blank(),
        axis.text    = element_text(family = "serif"))
}
```


\


## Provided data



The package `bridge` comes with two datasets:

- **Adour river**, an `sf` spatial object of type `LINESTRING` providing the 
geometry of the French river _L'Adour_;
- **Adour sites**, a `data.frame` with sites labels and sites coordinates along
the river.


\


### Adour river



Let's import the **Adour river** spatial layer.


```{r 'import-adour-river'}
# Import the spatial layer of Adour river ----
path_to_file <- system.file("extdata", "adour_lambert93.gpkg", package = "bridge")
adour_river  <- sf::st_read(path_to_file, quiet = TRUE)

head(adour_river)
```

This layer is an [`sf`](https://cran.r-project.org/package=sf) spatial object 
of type `LINESTRING` and contains one single row (geometry) defined in the 
[RGF93 / Lambert-93](https://epsg.io/2154) projection.


\


Now let's import a base map of France to locate the Adour river using the package
[`rnaturalearth`](https://cran.r-project.org/package=rnaturalearth).


```{r 'get-basemap'}
# Import France departments ----
france <- rnaturalearth::ne_states("france", returnclass = "sf")

# Remove overseas territories ----
france <- france[grep("^FR\\.", france$"code_hasc"), ]

# Project layer to RGF93 / Lambert-93 system ----
france <- sf::st_transform(france, sf::st_crs(adour_river))
```


\


Let's locate the Adour river of the France map.


```{r 'map-adour-river', fig.height=10, fig.width=10, out.width='80%'}
# Get extent of Adour river ----
adour_extent <- sf::st_bbox(adour_river)

# Add some margin around ----
adour_extent <- adour_extent + c(-10000, -10000, 10000, 10000)

ggplot() +
  geom_sf(data = france, fill = "lightgray", col = "white", size = 0.2) +
  geom_sf(data = adour_river, col = "steelblue") +
  geom_sf(data = sf::st_as_sfc(adour_extent), fill = NA, size = 0.3) +
  labs(title   = "The French river Adour", 
       caption = "RGF93 / Lambert-93 Projection") +
  custom_theme()
```


\


### Adour sites



Let's import the **Adour sites** data representing fictitious survey sites 
around the Adour.


```{r 'import-adour-sites'}
## Import the dataset of sites ----
path_to_file <- system.file("extdata", "adour_sites_coords.csv", package = "bridge")
adour_sites  <- read.csv(path_to_file)
adour_sites
```


This `data.frame` contains three columns: the site label and sites coordinates, also
defined in the [RGF93 / Lambert-93](https://epsg.io/2154) projection.



\


Let's convert this `data.frame` into an `sf` object of type `POINT`.

```{r 'df-to-sf', fig.height=10, fig.width=12}
# Convert data.frame to sf object ----
adour_sites_sf <- sf::st_as_sf(adour_sites, coords = 2:3, crs = "epsg:2154")
adour_sites_sf
```

**Note:** The package `bridge` expects that each site is one row in this object
and its geometry is `POINT` (other geometry like `MUTLIPOINT`, `POLYGON`, etc.)
are not implemented.


\


Finally, let's map these sites and the Adour river.


```{r 'map-adour-sites', fig.height=10, fig.width=12, out.width='80%'}
ggplot() +
  geom_sf(data = adour_river, col = "steelblue") +
  geom_sf(data = adour_sites_sf, shape = 19, size = 3) +
  geom_label(data = adour_sites, aes(x = longitude, y = latitude, label = site), 
            size = 4.2, nudge_x = 5000, family = "serif") +
  labs(title   = "Study area (French river Adour)", 
       caption = "RGF93 / Lambert-93 Projection") +
  custom_theme() +
  geom_segment(aes(x = 454180, xend = 440170, y = 6216290, yend = 6263320), 
               arrow = arrow(length = unit(0.75, 'cm'), type = 'closed'),
               size  = 2.25)
```

The arrow indicates the direction of the Adour river flow. This means that our 
system is a directed spatial network where the main direction is from upstream 
to downstream. 


\


## Nodes list

Let's identify the **nodes list**, i.e. a vector of unique **ordered** sites 
labels.


\


**Important:** as the sites have a direction (along the Adour river, from 
upstream to downstream), the order of the sites must be found in the sites labels.

For instance, in the Adour sites dataset, sites are labelled as **S-01**, **S-02**, 
..., **S-10**. This means that the most upstream site is **S-01** and the
most downstream is **S-10**.


\


But be aware that ordering characters can have some unwanted effects, especially if 
they contain numbers.

For instance,

```{r 'order-labels'}
# Alphanumerical ordering ----
sites <- c( "s1",  "s2", "s11", "s10")
sort(sites)

# Natural ordering ----
sites <- c("s01", "s02", "s11", "s10")
sort(sites)
```

You may want to code your sites as in the second example, i.e. based on a 
**Natural ordering**.


\


Let's use the function `nodes_list()` to 1) retrieve the unique labels of sites,
and to 2) order them along the Adour river (from upstream to downstream).


```{r 'nodes-list'}
adour_nodes <- nodes_list(adour_sites$"site")
adour_nodes
```


\


## Edges list


From this nodes list, we can create an **edges list**, i.e. a table describing 
links between nodes. Several methods exist to identify spatial neighbors and
most of them are available in the package 
[`spdep`](https://cran.r-project.org/package=spdep) (Bivand _et al._ 2022).


\


Here we implement our own method to detect spatial neighbors based sites labels 
and a degree of neighborhood. As our sites are ordered along the Adour river, if
we use a degree **1** of neighborhood, a site will be linked to the first next site 
(downstream) and also to the first previous site (upstream). If we use a degree 
**2** of neighborhood, a site will be linked to the two next sites 
(downstream) and also to the two previous sites (upstream). And so on.


\


Let's create this edges list with the function `edges_list()` provided by the package
`bridge`.

```{r 'edges-list'}
# List of edges with degree 1 of neighborhood ----
adour_edges <- edges_list(adour_nodes, degree = 1)
adour_edges
```

This `data.frame` contains four columns:

- `edge_id`: label of the edge
- `edge`: 0 (no edge) or 1 (edge), useful if `all = TRUE` (see below)
- `from`: label of the most upstream node of the edge
- `to`: label of the most downstream node of the edge

\


The function `edges_list()` has three additional arguments:

- `self`: if `TRUE`, a node can be linked to itself. Default is `FALSE`.
- `all`: if `TRUE`, the missing edges are also returned. Default is `FALSE`.
- `directed`: if `FALSE` (default), symmetrical edges (e.g. `S01-S02` and `S02-S01`) 
are returned (_undirected network_). Otherwise (_directed network_) only the 
first edge (e.g. `S01-S02`) is returned (according to the direction of the network).


\


**Tips: ** if you want to use Moran's Eigenvector Maps (MEM) use `directed = FALSE`.
If you want to use Asymetric Eigenvector Maps (AEM) use `directed = TRUE`.


\


Let's use this function `edges_list()` in a directed way.

```{r 'edges-list-directed'}
# List of edges with degree 1 of neighborhood (directed network) ----
adour_edges <- edges_list(adour_nodes, degree = 1, directed = TRUE)
adour_edges
```


\


The package `bridge` provides a convenient function to convert this edges list 
table to an `sf` object of type `LINESTRING`: `edges_to_sf()`.


\


Let's convert the edges list to a spatial object.

```{r 'edges-to-sf'}
# Convert edges list to sf object ----
adour_edges_sf <- edges_to_sf(adour_edges, adour_sites_sf)
adour_edges_sf
```


\


Let's add these edges to our map.


```{r 'map-adour-edges', fig.height=10, fig.width=12, out.width='80%'}
ggplot() +
  geom_sf(data = adour_river, col = "darkgray") +
  geom_sf(data = adour_edges_sf) +
  geom_sf(data = adour_sites_sf, shape = 19, size = 3) +
  labs(title   = "Edges with degree 1 of neighborhood", 
       caption = "RGF93 / Lambert-93 Projection") +
  custom_theme()
```


\


Finally, let's increase the degree of neighborhood.


```{r 'edges-degree-3'}
# List of edges with degree 3 of neighborhood ----
adour_edges <- edges_list(adour_nodes, degree = 3, directed = TRUE)

# Convert edges list to sf object ----
adour_edges_sf <- edges_to_sf(adour_edges, adour_sites_sf)
```

```{r 'map-adour-edges-3', fig.height=10, fig.width=12, out.width='80%'}
ggplot() +
  geom_sf(data = adour_river, col = "darkgray") +
  geom_sf(data = adour_edges_sf) +
  geom_sf(data = adour_sites_sf, shape = 19, size = 3) +
  labs(title   = "Edges with degree 3 of neighborhood", 
       caption = "RGF93 / Lambert-93 Projection") +
  custom_theme()
```


\


## Connectivity matrix


We can convert the edges list to a **connectivity matrix** (also called _adjacency matrix_) 
by using the function `adjacency_matrix()`. A connectivity matrix is a matrix of 
dimensions `n x n`, where `n` is the number of nodes indicating the presence of
an edge between two nodes (noted as `1`) or the absence of edge (noted as `0`).


\


**Note:** this object is useless to work with AEM, but can useful to check the edges
list.


\


The function `adjacency_matrix()` of the package `bridge` only requires the edges list as argument.

```{r 'connectivity-matrix'}
# Create list of edges ----
adour_edges <- edges_list(adour_nodes, degree = 1, directed = TRUE)
adour_edges

# Create connectivity matrix ----
adour_edges_mat <- adjacency_matrix(adour_edges)
adour_edges_mat
```

See the documentation of the function `adjacency_matrix()` for other usage.


\


The package `bridge` provides a convenient function to visualize matrix: 
`plot_matrix()`.


\


Let's plot our connectivity matrix.


```{r 'plot-connectivity-matrix', fig.height=10, fig.width=10, out.width='80%'}
plot_matrix(adour_edges_mat, title = "Connectivity matrix")
```


\


## Nodes-by-edges matrix

The first object required by AEM is the nodes-by-edges matrix, i.e. a matrix of 
dimensions `n x k`, where `n` is the number of nodes (in rows) and `k` the 
number of edges (in columns). This matrix represents the presence of a link 
influencing a site, directly or indirectly. It also gives the direction of the 
network.


\


The function `nodes_by_edges_matrix()` of the package `bridge` is an alternative
to the function [`aem.build.binary()`](https://sdray.github.io/adespatial/reference/aem.build.binary.html) of the package 
[`adespatial`](https://cran.r-project.org/package=adespatial). This function 
creates the same output but works in a different way: it's only based on nodes 
labels (not on spatial coordinates). In addition, the function `nodes_by_edges_matrix()` 
adds labels to nodes and edges.



```{r 'nodes-by-edges-matrix'}
# Create nodes-by-edges matrix ----
adour_edges_mat <- nodes_by_edges_matrix(adour_edges)
adour_edges_mat
```

This function returns a list with the following elements:

- `se.mat`: the nodes-by-edges matrix of dimensions `n x k`, where `n` is the 
number of nodes and `k` the number of edges (including the edge between the 
fictitious origin and the first site);
- `edges`: a `data.frame` of edges list.


\


**Note:** a fictitious `origin` site (coded as `0`) has been created and represents
the first edge (coded as `E-01`) to the first site.


\


Let's plot our connectivity matrix.


```{r 'plot-nodes-by-edges-matrix', fig.height=10, fig.width=10, out.width='80%'}
plot_matrix(adour_edges_mat$"se.mat", title = "Nodes-by-edges matrix")
```


\


## Edges weights


### Euclidean distance


The Euclidean distance, a straightforward metric to compute distances between 
pairs of nodes, is implemented in the function `distance_euclidean()`. This 
function only requires the coordinates of sites (object `sf`) and will compute 
distances between each **possible** pairs of sites.


```{r 'euclidean-distance'}
# Euclidean distance between pairs of sites ----
adour_dists <- distance_euclidean(adour_sites_sf)

head(adour_dists)
```

The returned object is a `data.frame` will the following three columns:

- `from`, the label of the first node
- `to`, the label of the second node
- `weight`, the Euclidean distance between two nodes


\


Note that if the units of the coordinates system are in degrees (geodetic 
coordinates), the Great Circle distance will be used instead. See the 
documentation of the function `distance_euclidean()` for further detail.


\


### Distance along river


The function `distance_along()` implements a more complex way to compute distances
between pairs of sites by taking into account the shape of the river. The distance
between two sites will be equal to the length of river segment between these two sites.

To create segments, this function will sample points on the river and select the two points
of the river that are the nearest from the two sites. These two points will define
the endpoints of the segment.

```{r 'distance-along', eval = FALSE}
# Distance along the river between pairs of sites ----
distance_along(adour_sites_sf, adour_river, density = 0.01, type = "regular")
```

This function uses two specific arguments:

- `density`: the density of points to sample on the linear structure. 
(see `sf::st_line_sample()` for further detail)
- `type`: the method to sample points on the linear shape. Either `regular` 
(default) or `random`


\


The returned object has the same format as in `distance_euclidean()`.


\


## Edges weights vector


The second object (optional) required by AEM is the **edges weights vector**. It is
a simple numeric vector giving the weight of each edge. Here we are going to use
the Euclidean distance as the edge weight.


\


The function `edges_weights_vector()` of the package `bridge` takes the output
of the function `distance_euclidean()` (or `distance_along()`) to create this
edges weights vector.


```{r 'edges-weights-vector'}
# Edges weights vector ----
adour_weights <- edges_weights_vector(adour_edges_mat, adour_dists)
adour_weights
```


\


Now we can run an AEM as follow: `adespatial::aem(adour_edges_mat, weight = adour_weights)`


\


**Note:** if you want to remove the first (fictitious) edge, you need to use the 
function as follow: `adespatial::aem(adour_edges_mat, weight = adour_weights[-1], rm.link0 = TRUE)`


\


## References

Bivand R. _et al._ (2022) spdep: Spatial dependence, weighting schemes, and statistics.
R package version 1.2-4, 
[https://CRAN.R-project.org/package=spdep](https://CRAN.R-project.org/package=spdep).

Blanchet F. G., Legendre P. & Borcard D. (2008) Modelling directional spatial 
processes in ecological data. **Ecological Modelling**, 215, 325-336. 
doi: [10.1016/j.ecolmodel.2008.04.001](https://doi.org/10.1016/j.ecolmodel.2008.04.001).

Dray S., Legendre P. & Peres-Neto P. R. (2006) Spatial modelling: a comprehensive 
framework for principal coordinate analysis of neighbour matrices (PCNM).
**Ecological Modelling**, 196, 483-493. 
doi: [10.1016/j.ecolmodel.2006.02.015](https://doi.org/10.1016/j.ecolmodel.2006.02.015).

Dray S. _et al._ (2022) adespatial: Multivariate Multiscale Spatial Analysis. 
R package version 0.3-16, 
[https://CRAN.R-project.org/package=adespatial](https://CRAN.R-project.org/package=adespatial).
