---
title: "Chess pieces"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Chess pieces}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: inline
---

```{r include=FALSE}
knitr::opts_chunk$set(collapse  = TRUE,
                      comment   = "#>",
                      out.width = "50%",
                      dpi       = 72,
                      fig.align = "center")
```


```{r 'setup', echo=FALSE}
library("chessboard")
library("ggplot2")
library("patchwork")
```


```{r 'fictitious-data', echo=FALSE}
sites <- expand.grid("transect" = 1:9, "quadrat" = 1:9)

nodes <- create_node_labels(data     = sites,
                            transect = "transect", 
                            quadrat  = "quadrat")

focus  <- "5-5"

transects_quadrats <- expand.grid("transect" = 1:3, 
                                  "quadrat"  = 1:5)

transects_only     <- data.frame("transect"  = 1:3)
quadrats_only      <- data.frame("quadrat"   = 1:5)

nodes_transects_quadrats <- create_node_labels(data     = transects_quadrats,
                                               transect = "transect", 
                                               quadrat  = "quadrat")

nodes_transects_only <- create_node_labels(data     = transects_only,
                                           transect = "transect")

nodes_quadrats_only  <- create_node_labels(data     = quadrats_only,
                                           quadrat  = "quadrat")
```




The package `chessboard` implements different methods for detecting neighbors. 
All are derived from the [chess game](https://en.wikipedia.org/wiki/Chess#Movement) 
and allow user to define complex moves to create direct edges (links) between 
nodes (sampling units). The following table provides a description of the chess 
pieces available in `chessboard`.


\


| Function       | 2D network | Transect only | Quadrat only | Description                                              |
| :------------- | :--------: | :-----------: | :----------: | :------------------------------------------------------- |
| `pawn()`       |     ✅     |      ❌       |      ✅      | Finds neighbors vertically (i.e. along a transect)       |
| `fool()`       |     ✅     |      ✅       |      ❌      | Finds neighbors horizontally (i.e. along a quadrat)      |
| `rook()`       |     ✅     |      ❌       |      ❌      | Finds neighbors vertically and horizontally              |
| `bishop()`^\*^ |     ✅     |      ❌       |      ❌      | Finds neighbors along the two diagonals                  |
| `knight()`^\*^ |     ✅     |      ❌       |      ❌      | Finds neighbors in 'L' shape                             |
| `queen()`      |     ✅     |      ❌       |      ❌      | Finds neighbors horizontally, vertically, and diagonally |
| `wizard()`     |     ✅     |      ❌       |      ❌      | Finds neighbors in all directions                        |


^\*^ These two chess pieces each have two variants: `bishop_left()`, `bishop_right()` and `knight_left()`, `knight_right()`.


\


## Pawn


> The `pawn` can detect neighbors vertically, i.e. among quadrats along a transect. 

The function `pawn()` can be used to detect neighbors (black dots in Fig. 1) of a specific node (red dot in Fig. 1).

```{r eval=FALSE, echo=TRUE}
pawn(nodes, focus = "5-5", degree = 4, ...)
```

User can change the default settings (Fig. 1A),
by adding directionality (`directed = TRUE`, Fig. 1B) and reversing the default
directionality (`directed = TRUE` and `reverse = TRUE`, Fig. 1C).


\


```{r 'cb-pawn', eval=TRUE, fig.height=4.3, fig.width=12, echo=FALSE, fig.cap="Figure 1. Moves of the **Pawn** (with degree = 4)", out.width='100%'}
pawn_1 <- 
  gg_chessboard(nodes, "A. Undirected network", "") + 
  geom_node(nodes, focus) +
  geom_neighbors(nodes, pawn(nodes, focus, degree = 4, directed = FALSE, 
                             reverse = FALSE))

pawn_2 <- 
  gg_chessboard(nodes, "B. Directed network", "") + 
  geom_node(nodes, focus) +
  geom_neighbors(nodes, pawn(nodes, focus, degree = 4, directed = TRUE, 
                             reverse = FALSE))

pawn_3 <- 
  gg_chessboard(nodes, "C. Directed network (reverse)", "") + 
  geom_node(nodes, focus) +
  geom_neighbors(nodes, pawn(nodes, focus, degree = 4, directed = TRUE, 
                             reverse = TRUE))

(pawn_1 | pawn_2 | pawn_3)
```


\

Figure 2 shows the connectivity matrix of the 9 x 9 network when neighbors of all nodes
are detected by the `pawn` method and with a degree of neighborhood of 4.

```{r eval=FALSE, echo=TRUE}
create_edge_list(nodes, method = "pawn", degree = 4, ...)
```


```{r 'cb-pawn-mat', eval=TRUE, fig.height=4.3, fig.width=12, echo=FALSE, fig.cap="Figure 2. Connectivity matrices of the **Pawn** (with degree = 4)", out.width='100%'}
nb_1 <- create_edge_list(nodes, 
                          method   = "pawn", 
                          degree   = 4, 
                          directed = FALSE, 
                          reverse  = FALSE)

pawn_1 <- gg_matrix(connectivity_matrix(nb_1), "A. Undirected network") +
  theme(axis.text = element_blank(), axis.text.x = element_blank(),
        plot.caption = ggplot2::element_text(family = "mono", size = 12, 
            face = "bold", hjust = 0.5))

nb_2 <- create_edge_list(nodes, 
                          method   = "pawn", 
                          degree   = 4, 
                          directed = TRUE, 
                          reverse  = FALSE)

pawn_2 <- gg_matrix(connectivity_matrix(nb_2), "B. Directed network") +
  theme(axis.text = element_blank(), axis.text.x = element_blank(),
        plot.caption = ggplot2::element_text(family = "mono", size = 12, 
            face = "bold", hjust = 0.5))

nb_3 <- create_edge_list(nodes, 
                          method   = "pawn", 
                          degree   = 4, 
                          directed = TRUE, 
                          reverse  = TRUE)

pawn_3 <- gg_matrix(connectivity_matrix(nb_3), "C. Directed network (reverse)") +
  theme(axis.text = element_blank(), axis.text.x = element_blank(),
        plot.caption = ggplot2::element_text(family = "mono", size = 12, 
            face = "bold", hjust = 0.5))

(pawn_1 | pawn_2 | pawn_3)
```


\


## Fool

> The `fool` can detect neighbors horizontally, i.e. among transects along a quadrat. 

The function `fool()` can be used to detect neighbors (black dots in Fig. 3) of a specific node (red dot in Fig. 3).

```{r eval=FALSE, echo=TRUE}
fool(nodes, focus = "5-5", degree = 4, ...)
```

User can change the default settings (Fig. 3A), by adding directionality 
(`directed = TRUE`, Fig. 3B) and reversing the default directionality 
(`directed = TRUE` and `reverse = TRUE`, Fig. 3C).


\


```{r 'cb-fool', eval=TRUE, fig.height=4.3, fig.width=12, echo=FALSE, fig.cap="Figure 3. Moves of the **Fool** (with degree = 4)", out.width='100%'}
fool_1 <- 
  gg_chessboard(nodes, "A. Undirected network", "") + 
  geom_node(nodes, focus) +
  geom_neighbors(nodes, fool(nodes, focus, degree = 4, directed = FALSE, 
                             reverse = FALSE))

fool_2 <- 
  gg_chessboard(nodes, "B. Directed network", "") + 
  geom_node(nodes, focus) +
  geom_neighbors(nodes, fool(nodes, focus, degree = 4, directed = TRUE, 
                             reverse = FALSE))

fool_3 <- 
  gg_chessboard(nodes, "C. Directed network (reverse)", "") + 
  geom_node(nodes, focus) +
  geom_neighbors(nodes, fool(nodes, focus, degree = 4, directed = TRUE, 
                             reverse = TRUE))

(fool_1 | fool_2 | fool_3)
```


\


Figure 4 shows the connectivity matrix of the 9 x 9 network when neighbors of all nodes
are detected by the `fool` method and with a degree of neighborhood of 4.

```{r eval=FALSE, echo=TRUE}
create_edge_list(nodes, method = "fool", degree = 4, ...)
```


\


```{r 'cb-fool-mat', eval=TRUE, fig.height=4.3, fig.width=12, echo=FALSE, fig.cap="Figure 4. Connectivity matrices of the **Fool** (with degree = 4)", out.width='100%'}
nb_1 <- create_edge_list(nodes, 
                          method   = "fool", 
                          degree   = 4, 
                          directed = FALSE, 
                          reverse  = FALSE)

fool_1 <- gg_matrix(connectivity_matrix(nb_1), "A. Undirected network") +
  theme(axis.text = element_blank(), axis.text.x = element_blank(),
        plot.caption = ggplot2::element_text(family = "mono", size = 12, 
            face = "bold", hjust = 0.5))

nb_2 <- create_edge_list(nodes, 
                          method   = "fool", 
                          degree   = 4, 
                          directed = TRUE, 
                          reverse  = FALSE)

fool_2 <- gg_matrix(connectivity_matrix(nb_2), "B. Directed network") +
  theme(axis.text = element_blank(), axis.text.x = element_blank(),
        plot.caption = ggplot2::element_text(family = "mono", size = 12, 
            face = "bold", hjust = 0.5))

nb_3 <- create_edge_list(nodes, 
                          method   = "fool", 
                          degree   = 4, 
                          directed = TRUE, 
                          reverse  = TRUE)

fool_3 <- gg_matrix(connectivity_matrix(nb_3), "C. Directed network (reverse)") +
  theme(axis.text = element_blank(), axis.text.x = element_blank(),
        plot.caption = ggplot2::element_text(family = "mono", size = 12, 
            face = "bold", hjust = 0.5))

(fool_1 | fool_2 | fool_3)
```



## Rook


> The `rook` can detect neighbors both horizontally and vertically, 
i.e. among transects along a quadrat and among quadrats along a transect. 

The function `rook()` can be used to detect neighbors (black dots in Fig. 5) of a specific node (red dot in Fig. 5).

```{r eval=FALSE, echo=TRUE}
rook(nodes, focus = "5-5", degree = 4, ...)
```

User can change the default settings (Fig. 5A), by adding directionality 
(`directed = TRUE`, Fig. 5B) and reversing the default directionality 
(`directed = TRUE` and `reverse = TRUE`, Fig. 5C).


\


```{r 'cb-rook', eval=TRUE, fig.height=4.3, fig.width=12, echo=FALSE, fig.cap="Figure 5. Moves of the **Rook** (with degree = 4)", out.width='100%'}
rook_1 <- 
  gg_chessboard(nodes, "A. Undirected network", "") + 
  geom_node(nodes, focus) +
  geom_neighbors(nodes, rook(nodes, focus, degree = 4, directed = FALSE, 
                             reverse = FALSE))

rook_2 <- 
  gg_chessboard(nodes, "B. Directed network", "") + 
  geom_node(nodes, focus) +
  geom_neighbors(nodes, rook(nodes, focus, degree = 4, directed = TRUE, 
                             reverse = FALSE))

rook_3 <- 
  gg_chessboard(nodes, "C. Directed network (reverse)", "") + 
  geom_node(nodes, focus) +
  geom_neighbors(nodes, rook(nodes, focus, degree = 4, directed = TRUE, 
                             reverse = TRUE))

(rook_1 | rook_2 | rook_3)
```


\


Figure 6 shows the connectivity matrix of the 9 x 9 network when neighbors of all nodes
are detected by the `rook` method and with a degree of neighborhood of 4.

```{r eval=FALSE, echo=TRUE}
create_edge_list(nodes, method = "rook", degree = 4, ...)
```


\


```{r 'cb-rook-mat', eval=TRUE, fig.height=4.3, fig.width=12, echo=FALSE, fig.cap="Figure 6. Connectivity matrices of the **Rook** (with degree = 4)", out.width='100%'}
nb_1 <- create_edge_list(nodes, 
                          method   = "rook", 
                          degree   = 4, 
                          directed = FALSE, 
                          reverse  = FALSE)

rook_1 <- gg_matrix(connectivity_matrix(nb_1), "A. Undirected network") +
  theme(axis.text = element_blank(), axis.text.x = element_blank(),
        plot.caption = ggplot2::element_text(family = "mono", size = 12, 
            face = "bold", hjust = 0.5))

nb_2 <- create_edge_list(nodes, 
                          method   = "rook", 
                          degree   = 4, 
                          directed = TRUE, 
                          reverse  = FALSE)

rook_2 <- gg_matrix(connectivity_matrix(nb_2), "B. Directed network") +
  theme(axis.text = element_blank(), axis.text.x = element_blank(),
        plot.caption = ggplot2::element_text(family = "mono", size = 12, 
            face = "bold", hjust = 0.5))

nb_3 <- create_edge_list(nodes, 
                          method   = "rook", 
                          degree   = 4, 
                          directed = TRUE, 
                          reverse  = TRUE)

rook_3 <- gg_matrix(connectivity_matrix(nb_3), "C. Directed network (reverse)") +
  theme(axis.text = element_blank(), axis.text.x = element_blank(),
        plot.caption = ggplot2::element_text(family = "mono", size = 12, 
            face = "bold", hjust = 0.5))

(rook_1 | rook_2 | rook_3)
```



\


## Bishop


> The `bishop` can detect neighbors diagonally. 

The function `bishop()` can be used to detect neighbors (black dots in Fig. 7) of a specific node (red dot in Fig. 7).

```{r eval=FALSE, echo=TRUE}
bishop(nodes, focus = "5-5", degree = 4, ...)
```

User can change the default settings (Fig. 7A), by adding directionality 
(`directed = TRUE`, Fig. 7B) and reversing the default directionality 
(`directed = TRUE` and `reverse = TRUE`, Fig. 7C).


\


```{r 'cb-bishop', eval=TRUE, fig.height=4.3, fig.width=12, echo=FALSE, fig.cap="Figure 7. Moves of the **Bishop** (with degree = 4)", out.width='100%'}
bishop_1 <- 
  gg_chessboard(nodes, "A. Undirected network", "") + 
  geom_node(nodes, focus) +
  geom_neighbors(nodes, bishop(nodes, focus, degree = 4, directed = FALSE, 
                               reverse = FALSE))

bishop_2 <- 
  gg_chessboard(nodes, "B. Directed network", "") + 
  geom_node(nodes, focus) +
  geom_neighbors(nodes, bishop(nodes, focus, degree = 4, directed = TRUE, 
                               reverse = FALSE))

bishop_3 <- 
  gg_chessboard(nodes, "C. Directed network (reverse)", "") + 
  geom_node(nodes, focus) +
  geom_neighbors(nodes, bishop(nodes, focus, degree = 4, directed = TRUE, 
                               reverse = TRUE))

(bishop_1 | bishop_2 | bishop_3)
```


\


Figure 8 shows the connectivity matrix of the 9 x 9 network when neighbors of all nodes
are detected by the `bishop` method and with a degree of neighborhood of 4.

```{r eval=FALSE, echo=TRUE}
create_edge_list(nodes, method = "bishop", degree = 4, ...)
```


\


```{r 'cb-bishop-mat', eval=TRUE, fig.height=4.3, fig.width=12, echo=FALSE, fig.cap="Figure 8. Connectivity matrices of the **Bishop** (with degree = 4)", out.width='100%'}
nb_1 <- create_edge_list(nodes, 
                          method   = "bishop", 
                          degree   = 4, 
                          directed = FALSE, 
                          reverse  = FALSE)

bishop_1 <- gg_matrix(connectivity_matrix(nb_1), "A. Undirected network") +
  theme(axis.text = element_blank(), axis.text.x = element_blank(),
        plot.caption = ggplot2::element_text(family = "mono", size = 12, 
            face = "bold", hjust = 0.5))

nb_2 <- create_edge_list(nodes, 
                          method   = "bishop", 
                          degree   = 4, 
                          directed = TRUE, 
                          reverse  = FALSE)

bishop_2 <- gg_matrix(connectivity_matrix(nb_2), "B. Directed network") +
  theme(axis.text = element_blank(), axis.text.x = element_blank(),
        plot.caption = ggplot2::element_text(family = "mono", size = 12, 
            face = "bold", hjust = 0.5))

nb_3 <- create_edge_list(nodes, 
                          method   = "bishop", 
                          degree   = 4, 
                          directed = TRUE, 
                          reverse  = TRUE)

bishop_3 <- gg_matrix(connectivity_matrix(nb_3), "C. Directed network (reverse)") +
  theme(axis.text = element_blank(), axis.text.x = element_blank(),
        plot.caption = ggplot2::element_text(family = "mono", size = 12, 
            face = "bold", hjust = 0.5))

(bishop_1 | bishop_2 | bishop_3)
```



\


## Knight

> The `knight` has a complex move, like an 'L' shape. It's the difference between
the `wizard` and the `queen`.

The function `knight()` can be used to detect neighbors (black dots in Fig. 9) of a specific node (red dot in Fig. 9).

```{r eval=FALSE, echo=TRUE}
knight(nodes, focus = "5-5", degree = 4, ...)
```

User can change the default settings (Fig. 9A), by adding directionality 
(`directed = TRUE`, Fig. 9B) and reversing the default directionality 
(`directed = TRUE` and `reverse = TRUE`, Fig. 9C).

\


```{r 'cb-knight', eval=TRUE, fig.height=4.3, fig.width=12, echo=FALSE, fig.cap="Figure 9. Moves of the **Knight** (with degree = 4)", out.width='100%'}
knight_1 <- 
  gg_chessboard(nodes, "A. Undirected network", "") + 
  geom_node(nodes, focus) +
  geom_neighbors(nodes, knight(nodes, focus, degree = 4, directed = FALSE, 
                               reverse = FALSE))

knight_2 <- 
  gg_chessboard(nodes, "B. Directed network", "") + 
  geom_node(nodes, focus) +
  geom_neighbors(nodes, knight(nodes, focus, degree = 4, directed = TRUE, 
                               reverse = FALSE))

knight_3 <- 
  gg_chessboard(nodes, "C. Directed network (reverse)", "") + 
  geom_node(nodes, focus) +
  geom_neighbors(nodes, knight(nodes, focus, degree = 4, directed = TRUE, 
                               reverse = TRUE))

(knight_1 | knight_2 | knight_3)
```


\


Figure 10 shows the connectivity matrix of the 9 x 9 network when neighbors of all nodes
are detected by the `knight` method and with a degree of neighborhood of 4.

```{r eval=FALSE, echo=TRUE}
create_edge_list(nodes, method = "knight", degree = 4, ...)
```


\


```{r 'cb-knight-mat', eval=TRUE, fig.height=4.3, fig.width=12, echo=FALSE, fig.cap="Figure 10. Connectivity matrices of the **Knight** (with degree = 4)", out.width='100%'}
nb_1 <- create_edge_list(nodes, 
                          method   = "knight", 
                          degree   = 4, 
                          directed = FALSE, 
                          reverse  = FALSE)

knight_1 <- gg_matrix(connectivity_matrix(nb_1), "A. Undirected network") +
  theme(axis.text = element_blank(), axis.text.x = element_blank(),
        plot.caption = ggplot2::element_text(family = "mono", size = 12, 
            face = "bold", hjust = 0.5))

nb_2 <- create_edge_list(nodes, 
                          method   = "knight", 
                          degree   = 4, 
                          directed = TRUE, 
                          reverse  = FALSE)

knight_2 <- gg_matrix(connectivity_matrix(nb_2), "B. Directed network") +
  theme(axis.text = element_blank(), axis.text.x = element_blank(),
        plot.caption = ggplot2::element_text(family = "mono", size = 12, 
            face = "bold", hjust = 0.5))

nb_3 <- create_edge_list(nodes, 
                          method   = "knight", 
                          degree   = 4, 
                          directed = TRUE, 
                          reverse  = TRUE)

knight_3 <- gg_matrix(connectivity_matrix(nb_3), "C. Directed network (reverse)") +
  theme(axis.text = element_blank(), axis.text.x = element_blank(),
        plot.caption = ggplot2::element_text(family = "mono", size = 12, 
            face = "bold", hjust = 0.5))

(knight_1 | knight_2 | knight_3)
```



\


## Queen


> The `queen` can detect neighbors horizontally, vertically and 
diagonally. It's a combination the `rook` and the `bishop`.

The function `queen()` can be used to detect neighbors (black dots in Fig. 11) of a specific node (red dot in Fig. 11).

```{r eval=FALSE, echo=TRUE}
queen(nodes, focus = "5-5", degree = 4, ...)
```

User can change the default settings (Fig. 11A), by adding directionality 
(`directed = TRUE`, Fig. 11B) and reversing the default directionality 
(`directed = TRUE` and `reverse = TRUE`, Fig. 11C).

\


```{r 'cb-queen', eval=TRUE, fig.height=4.3, fig.width=12, echo=FALSE, fig.cap="Figure 11. Moves of the **Queen** (with degree = 4)", out.width='100%'}
queen_1 <- 
  gg_chessboard(nodes, "A. Undirected network", "") + 
  geom_node(nodes, focus) +
  geom_neighbors(nodes, queen(nodes, focus, degree = 4, directed = FALSE, 
                              reverse = FALSE))

queen_2 <- 
  gg_chessboard(nodes, "B. Directed network", "") + 
  geom_node(nodes, focus) +
  geom_neighbors(nodes, queen(nodes, focus, degree = 4, directed = TRUE, 
                              reverse = FALSE))

queen_3 <- 
  gg_chessboard(nodes, "C. Directed network (reverse)", "") + 
  geom_node(nodes, focus) +
  geom_neighbors(nodes, queen(nodes, focus, degree = 4, directed = TRUE, 
                              reverse = TRUE))

(queen_1 | queen_2 | queen_3)
```


\


Figure 12 shows the connectivity matrix of the 9 x 9 network when neighbors of all nodes
are detected by the `queen` method and with a degree of neighborhood of 4.

```{r eval=FALSE, echo=TRUE}
create_edge_list(nodes, method = "queen", degree = 4, ...)
```


\


```{r 'cb-queen-mat', eval=TRUE, fig.height=4.3, fig.width=12, echo=FALSE, fig.cap="Figure 12. Connectivity matrices of the **Queen** (with degree = 4)", out.width='100%'}
nb_1 <- create_edge_list(nodes, 
                          method   = "queen", 
                          degree   = 4, 
                          directed = FALSE, 
                          reverse  = FALSE)

queen_1 <- gg_matrix(connectivity_matrix(nb_1), "A. Undirected network") +
  theme(axis.text = element_blank(), axis.text.x = element_blank(),
        plot.caption = ggplot2::element_text(family = "mono", size = 12, 
            face = "bold", hjust = 0.5))

nb_2 <- create_edge_list(nodes, 
                          method   = "queen", 
                          degree   = 4, 
                          directed = TRUE, 
                          reverse  = FALSE)

queen_2 <- gg_matrix(connectivity_matrix(nb_2), "B. Directed network") +
  theme(axis.text = element_blank(), axis.text.x = element_blank(),
        plot.caption = ggplot2::element_text(family = "mono", size = 12, 
            face = "bold", hjust = 0.5))

nb_3 <- create_edge_list(nodes, 
                          method   = "queen", 
                          degree   = 4, 
                          directed = TRUE, 
                          reverse  = TRUE)

queen_3 <- gg_matrix(connectivity_matrix(nb_3), "C. Directed network (reverse)") +
  theme(axis.text = element_blank(), axis.text.x = element_blank(),
        plot.caption = ggplot2::element_text(family = "mono", size = 12, 
            face = "bold", hjust = 0.5))

(queen_1 | queen_2 | queen_3)
```


\


## Wizard

> The `wizard` can detect neighbors in any direction. It's a combination the 
`queen` and the `knight`.

The function `wizard()` can be used to detect neighbors (black dots in Fig. 13) of a specific node (red dot in Fig. 13).

```{r eval=FALSE, echo=TRUE}
wizard(nodes, focus = "5-5", degree = 4, ...)
```

User can change the default settings (Fig. 13A), by adding directionality 
(`directed = TRUE`, Fig. 13B) and reversing the default directionality 
(`directed = TRUE` and `reverse = TRUE`, Fig. 13C).
\


```{r 'cb-wizard', eval=TRUE, fig.height=4.3, fig.width=12, echo=FALSE, fig.cap="Figure 13. Moves of the **Wizard** (with degree = 4)", out.width='100%'}
wizard_1 <- 
  gg_chessboard(nodes, "A. Undirected network", "") + 
  geom_node(nodes, focus) +
  geom_neighbors(nodes, wizard(nodes, focus, degree = 4, directed = FALSE, 
                               reverse = FALSE))

wizard_2 <- 
  gg_chessboard(nodes, "B. Directed network", "") + 
  geom_node(nodes, focus) +
  geom_neighbors(nodes, wizard(nodes, focus, degree = 4, directed = TRUE, 
                               reverse = FALSE))

wizard_3 <- 
  gg_chessboard(nodes, "C. Directed network (reverse)", "") + 
  geom_node(nodes, focus) +
  geom_neighbors(nodes, wizard(nodes, focus, degree = 4, directed = TRUE, 
                               reverse = TRUE))

(wizard_1 | wizard_2 | wizard_3)
```


\


Figure 14 shows the connectivity matrix of the 9 x 9 network when neighbors of all nodes
are detected by the `wizard` method and with a degree of neighborhood of 4.

```{r eval=FALSE, echo=TRUE}
create_edge_list(nodes, method = "wizard", degree = 4, ...)
```


\


```{r 'cb-wizard-mat', eval=TRUE, fig.height=4.3, fig.width=12, echo=FALSE, fig.cap="Figure 14. Connectivity matrices of the **Wizard** (with degree = 4)", out.width='100%'}
nb_1 <- create_edge_list(nodes, 
                          method   = "wizard", 
                          degree   = 4, 
                          directed = FALSE, 
                          reverse  = FALSE)

wizard_1 <- gg_matrix(connectivity_matrix(nb_1), "A. Undirected network") +
  theme(axis.text = element_blank(), axis.text.x = element_blank(),
        plot.caption = ggplot2::element_text(family = "mono", size = 12, 
            face = "bold", hjust = 0.5))

nb_2 <- create_edge_list(nodes, 
                          method   = "wizard", 
                          degree   = 4, 
                          directed = TRUE, 
                          reverse  = FALSE)

wizard_2 <- gg_matrix(connectivity_matrix(nb_2), "B. Directed network") +
  theme(axis.text = element_blank(), axis.text.x = element_blank(),
        plot.caption = ggplot2::element_text(family = "mono", size = 12, 
            face = "bold", hjust = 0.5))

nb_3 <- create_edge_list(nodes, 
                          method   = "wizard", 
                          degree   = 4, 
                          directed = TRUE, 
                          reverse  = TRUE)

wizard_3 <- gg_matrix(connectivity_matrix(nb_3), "C. Directed network (reverse)") +
  theme(axis.text = element_blank(), axis.text.x = element_blank(),
        plot.caption = ggplot2::element_text(family = "mono", size = 12, 
            face = "bold", hjust = 0.5))

(wizard_1 | wizard_2 | wizard_3)
```


\


## Custom moves


It's possible to create any kind of scenarios by combining the previous moves
and by changing the values of the arguments `degree`, `directed`, and `reverse`.


```{r 'cb-custom', eval=TRUE, fig.height=4.3, fig.width=12, echo=FALSE, fig.cap="Figure 15. Custom moves", out.width='100%'}
wizard_1 <- 
  gg_chessboard(nodes, "Pawn & Bishop", "") + 
  geom_node(nodes, focus) +
  geom_neighbors(nodes, pawn(nodes, focus, degree = 3, directed = TRUE, 
                             reverse = FALSE)) +
  geom_neighbors(nodes, bishop(nodes, focus, degree = 4, directed = TRUE, 
                               reverse = FALSE))

wizard_2 <- 
  gg_chessboard(nodes, "Knight L & Bishop L", "") + 
  geom_node(nodes, focus) +
  geom_neighbors(nodes, knight_left(nodes, focus, degree = 4, directed = TRUE, 
                                    reverse = FALSE)) +
  geom_neighbors(nodes, bishop_left(nodes, focus, degree = 4, directed = TRUE, 
                                    reverse = FALSE))

wizard_3 <- 
  gg_chessboard(nodes, "Knight L, Bishop L, Pawn & Fool", "") + 
  geom_node(nodes, focus) +
  geom_neighbors(nodes, knight_left(nodes, focus, degree = 4, directed = TRUE, 
                                    reverse = FALSE)) +
  geom_neighbors(nodes, bishop_left(nodes, focus, degree = 4, directed = TRUE, 
                                    reverse = FALSE)) +
  geom_neighbors(nodes, pawn(nodes, focus, degree = 4, directed = TRUE, 
                             reverse = FALSE)) +
  geom_neighbors(nodes, fool(nodes, focus, degree = 4, directed = TRUE, 
                             reverse = TRUE))

(wizard_1 | wizard_2 | wizard_3)
```
