---
title: "Get started"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Get started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r include=FALSE}
knitr::opts_chunk$set(collapse  = TRUE,
                      comment   = "#>",
                      out.width = "100%",
                      dpi       = 96,
                      fig.align = "center")
```


The aim of the package `chessboard` is to provide tools to work with **directed**
(asymmetric) and **undirected** (symmetrical) spatial **networks**.
It makes easier the creation of **connectivity matrix**, i.e. a binary matrix 
of dimensions `n x n`, where `n` is the number of **nodes** (sampling units) 
indicating the presence (`1`) or the absence (`0`) of an **edge** (link) 
between pairs of nodes. Different network objects can be produced by `chessboard`:

- **nodes list**
- **neighbors list**
- **edges list**
- **connectivity matrix**


\


In addition, the package can also produce objects that can be used later in 
Asymetric Eigenvector Maps (AEM, Blanchet _et al._ 2008), method available in 
the package [`adespatial`](https://cran.r-project.org/package=adespatial) (Dray _et al._ 2022):
**nodes by edges matrix** and **edges weights vector**.


\


This package has been developed for the 
[FRB-CESAB](https://www.fondationbiodiversite.fr/en/about-the-foundation/le-cesab/) 
working group 
[Bridge](https://www.fondationbiodiversite.fr/en/the-frb-in-action/programs-and-projects/le-cesab/bridge/) 
that aims to better understand the role of local and regional environmental 
factors in shaping the taxonomic and functional diversity of plant communities 
established along river corridors, roadside corridors and cultivated field 
margins.


\


**Setup**



```{r 'setup'}
library("chessboard")
library("ggplot2")
library("patchwork")
```



```{r 'ggplot-theme', echo=FALSE}
## Custom ggplot2 theme ----
custom_theme <- function() {
  theme_light() + 
  theme(plot.title   = element_text(face = "bold", family = "serif", size = 18),
        plot.caption = element_text(face = "italic", family = "serif"),
        axis.title   = element_blank(),
        axis.text    = element_text(family = "serif"))
}
```


**N.B.** The package `chessboard` can also deal with one-dimensional sampling 
(either **transects-only** or **quadrats-only**) and with **undirected** networks.


\


## Nodes labeling


It's **important** to note that, even the package `chessboard` can handle spatial 
networks, it does not explicitly use spatial coordinates to find neighbors (and edges). 
Instead, it will identify neighbors only based on **nodes labels**. In other 
words, the order of nodes labels will give the dimension and the direction of 
the network.


\

Let's import the survey dataset provided by the package.

```{r 'import-adour-survey'}
## Import survey data ----
path_to_file <- system.file("extdata", "adour_survey_sampling.csv",
                            package = "chessboard")
adour_sites  <- read.csv(path_to_file)

head(adour_sites, 20)
```

This `data.frame` contains the following columns: 

- `location`: the identifier of the location (`numeric`)
- `transect`: the identifier of the transect (`numeric`)
- `quadrat`: the identifier of the quadrat (`numeric`)
- `longitude`: the longitude of the site (**node**) defined in the [RGF93 / Lambert-93](https://epsg.io/2154) projection
- `latitude`: the latitude of the site (**node**) defined in the [RGF93 / Lambert-93](https://epsg.io/2154) projection


\


**N.B.** The column `location` is optional (if the survey has been conducted at one single location).
If the network has one dimension, one of the columns `transect` and `quadrat` can
be omitted.


\



\

From this figure, several things should be noted:

1. For two-dimensional networks, the x-axis **always** corresponds to transects
and the y-axis to the quadrats.
2. In this survey, the locations have the same transect identifiers (1 to 3) 
but the **identifier of quadrats increases** with locations.
3. The **order of the quadrats** follows the direction of the network: the first 
quadrat is upstream and the last is downstream. 


\


When using the package `chessboard`, the **first step** is to create **nodes labels**
by combining the identifiers of transects and quadrats. It's the purpose of the
function `create_node_labels()`.


\

But first, let's select the first location (optional).


```{r 'select-location'}
## Select first location (optional) ----
adour_sites <- adour_sites[adour_sites$"location" == 1, ]
adour_sites
```


\


Now, we can create the nodes labels.

```{r 'nodes-labels'}
## Create nodes labels ----
adour_sites <- create_node_labels(data     = adour_sites, 
                                  location = "location",
                                  transect = "transect", 
                                  quadrat  = "quadrat")

adour_sites
```

\

The nodes labels are of the form: `1-2`, where 
`1` is the identifier of the **transect** (created by the function if missing, 
i.e. quadrats-only sampling), and 
`2`, the identifier of the **quadrat** (created by the function if missing,
i.e. transects-only sampling).


\

The function `get_nodes_list()` can be used to extract the nodes list.

```{r 'get-nodes-labels'}
## Get nodes labels ----
get_nodes_list(adour_sites)
```



\


## Edges list



The creation of a list of edges (links) between nodes (sampling units) is based 
on the detection of neighbors. The function `create_edges_list()` implements 
three neighborhood rules:

1. **Degree of neighborhood** (argument `degree`): the number of adjacent 
nodes that will be used to create **direct** edges. If `degree = 1`, 
only nodes directly adjacent to the focal node will be considered as 
neighbors.
2. **Orientation of neighborhood** (argument `method`): can neighbors be 
detecting horizontally and/or vertically and/or diagonally? The package 
`chessboard` implements all possible orientations derived from the chess 
game.
3. **Direction of neighborhood** (arguments `directed` and `reverse`): does 
the sampling design has a direction? If so, the network 
will be considered as **directed** and the direction will follow the order 
of nodes labels in both axes of the grid.


\


Different methods have implemented to define neighborhood (argument `method` 
of the function `create_edges_list()`). These methods
are named `'pawn'`, `'rook'`, `'bishop'`, `'queen'`, etc. A complete list of
available methods is available at: 
[https://frbcesab.github.io/chessboard/reference/index.html#detect-neighbors](https://frbcesab.github.io/chessboard/reference/index.html#detect-neighbors)


\


Before using the function `create_edges_list()`, users can explore these different
methods by calling the functions `pawn()`, `rook()`, `bishop()`, `queen()`, etc.

Let's take a look of the neighborhood method `pawn()`.

```{r 'method-pawn'}
## Explore pawn method to find neighbors ----
neighbors_pawn <- pawn(nodes    = adour_sites, 
                       focus    = "2-3", 
                       degree   = 1, 
                       directed = FALSE, 
                       reverse  = FALSE)
neighbors_pawn
```

This function (and others) only works on a specific node (argument `focus`).


\


The package `chessboard` contains functions to visualize detected neighbors on a
chessboard: `gg_chessboard()` is used to plot a chessboard (dimensions defined by the
nodes list), `geom_node()` emphasizes the focus node (in red), and `geom_neighbors()` adds
the detected neighbors.


```{r 'nb-pawn', fig.height=8, fig.width=5, echo=FALSE, fig.cap="Figure 4. Detected neighbors (pawn method)", out.width='50%'}
gg_chessboard(adour_sites) +
  geom_node(adour_sites, focus = "2-3") +
  geom_neighbors(adour_sites, neighbors_pawn)
```


\


The Figure 5 shows the usage of the function `pawn()` (and others). 
This function can detect neighbors horizontally, i.e. among 
quadrats along a transect. User can change the default settings 
by increasing the degree of neighborhood (`degree = 4`, Fig. 5A),
by adding directionality (`directed = TRUE`, Fig. 5B), and 
by reversing the default directionality (`directed = TRUE` and `reverse = TRUE`, Fig. 5C).


```{r 'cb-pawn', fig.height=4.3, fig.width=12, echo=FALSE, fig.cap="Figure 5. Pawn movements", out.width='100%'}
demo_sites <- expand.grid("transect" = 1:9, "quadrat" = 1:9)

demo_nodes <- create_node_labels(data     = demo_sites,
                                 transect = "transect", 
                                 quadrat  = "quadrat")

demo_focus  <- "5-5"

pawn_1 <- 
  gg_chessboard(demo_nodes, "A. Undirected network", "") + 
  geom_node(demo_nodes, demo_focus) +
  geom_neighbors(demo_nodes, pawn(demo_nodes, demo_focus, degree = 4, 
                                  directed = FALSE, reverse = FALSE))

pawn_2 <- 
  gg_chessboard(demo_nodes, "B. Directed network", "") + 
  geom_node(demo_nodes, demo_focus) +
  geom_neighbors(demo_nodes, pawn(demo_nodes, demo_focus, degree = 4, 
                                  directed = TRUE, reverse = FALSE))

pawn_3 <- 
  gg_chessboard(demo_nodes, "C. Directed network (reverse)", "") + 
  geom_node(demo_nodes, demo_focus) +
  geom_neighbors(demo_nodes, pawn(demo_nodes, demo_focus, degree = 4, 
                                  directed = TRUE, reverse = TRUE))

(pawn_1 | pawn_2 | pawn_3)
```


\


Let's take another example with `method = "bishop"`. 
The function `bishop()` can detect neighbors diagonally. 
User can change the default settings 
by increasing the degree of neighborhood (`degree = 4`, Fig. 6A),
by adding directionality (`directed = TRUE`, Fig. 6B), and 
by reversing the default directionality (`directed = TRUE` and `reverse = TRUE`, Fig. 6C).


```{r 'cb-bishop', eval=TRUE, fig.height=4.3, fig.width=12, echo=FALSE, fig.cap="Figure 6. Bishop movements", out.width='100%'}
demo_sites <- expand.grid("transect" = 1:9, "quadrat" = 1:9)

demo_nodes <- create_node_labels(data     = demo_sites,
                                 transect = "transect", 
                                 quadrat  = "quadrat")

demo_focus  <- "5-5"

bishop_1 <- 
  gg_chessboard(demo_nodes, "A. Undirected network", "") + 
  geom_node(demo_nodes, demo_focus) +
  geom_neighbors(demo_nodes, bishop(demo_nodes, demo_focus, degree = 4, 
                                    directed = FALSE, reverse = FALSE))

bishop_2 <- 
  gg_chessboard(demo_nodes, "B. Directed network", "") + 
  geom_node(demo_nodes, demo_focus) +
  geom_neighbors(demo_nodes, bishop(demo_nodes, demo_focus, degree = 4, 
                                    directed = TRUE, reverse = FALSE))

bishop_3 <- 
  gg_chessboard(demo_nodes, "C. Directed network (reverse)", "") + 
  geom_node(demo_nodes, demo_focus) +
  geom_neighbors(demo_nodes, bishop(demo_nodes, demo_focus, degree = 4, 
                                    directed = TRUE, reverse = TRUE))

(bishop_1 | bishop_2 | bishop_3)
```


\

The vignette [Chess movements](https://frbcesab.github.io/chessboard/articles/movements.html) 
shows all possible movements implemented in `chessboard`.


\


Now, let's use the function `create_edges_list()` to create an edges list using the
method `'pawn'` with a degree `1` of neighborhood and in a directional way.


```{r 'create-edges-list-pawn'}
## Create edges list ----
adour_edges_pawn <- create_edges_list(nodes    = adour_sites, 
                                      method   = "pawn", 
                                      degree   = 1, 
                                      directed = TRUE,
                                      reverse  = FALSE,
                                      self     = FALSE)

adour_edges_pawn
```


\


It's possible to visualize these edges on a map, i.e. by using spatial coordinates.
First, we need to convert our sites into a spatial object (`POINT`).

```{r 'df-to-sf'}
## Convert sites to sf object ----
adour_sites_sf <- sf::st_as_sf(adour_sites, coords = c("longitude", "latitude"),
                               crs = "epsg:2154")

head(adour_sites_sf)
```


\


Now we can use the function `edges_to_sf()` to convert our edges list 
into a spatial object (`LINESTRING`).


```{r 'edges-list-to-sf'}
## Convert edges list to sf ----
adour_edges_pawn_sf <- edges_to_sf(edges = adour_edges_pawn, 
                                   sites = adour_sites_sf)

adour_edges_pawn_sf
```


\


We can now map our nodes and edges.

```{r 'map-edges-list-pawn', fig.height=8, fig.width=6.5, echo=TRUE, fig.cap="Figure 7. Edges list (pawn method)", out.width='80%'}
## Map of nodes and edges ----
ggplot(adour_sites_sf) +
  geom_sf(size = 12) +
  geom_sf(data = adour_edges_pawn_sf) +
  theme_light()
```


\


Users may want to combine different methods to detect neighbors. It's possible
by using for each method the function `create_edges_list()` and using the
function `append_edges_lists()` to merge all edges in a single list.


```{r 'create-edges-list-bishop'}
## Create edges list (Bishop method) ----
adour_edges_bishop <- create_edges_list(nodes    = adour_sites, 
                                        method   = "bishop", 
                                        degree   = 1, 
                                        directed = TRUE,
                                        reverse  = FALSE,
                                        self     = FALSE)

adour_edges_bishop

## Merge Pawn and Bishop edges ----
adour_edges <- append_edges_lists(adour_edges_pawn, adour_edges_bishop)

## Convert edges to spatial layer ----
adour_edges_sf <- edges_to_sf(adour_edges, adour_sites_sf)
```


\


```{r 'map-edges-list-pawn-bishop', fig.height=8, fig.width=6.5, echo=TRUE, fig.cap="Figure 8. Edges list (combined methods)", out.width='80%'}
## Map of nodes and edges ----
ggplot(adour_sites_sf) +
  geom_sf(size = 12) +
  geom_sf(data = adour_edges_sf) +
  theme_light()
```


\


## Connectivity matrix

From this edges list, we can build a **connectivity matrix**, i.e. a binary 
matrix of dimensions `n x n`, where `n` is the number of nodes (sampling units) 
indicating the presence (`1`) or the absence (`0`) of an edge (link) between pairs 
of nodes.

\

We can use the function `connectivity_matrix()` of the package `chessboard`.

```{r 'connectivity-matrix'}
## Create connectivity matrix ----
adour_matrix <- connectivity_matrix(adour_edges)

adour_matrix
```


\

The package `chessboard` provides a graphical function to visualize this matrix:
`gg_matrix()`.

```{r 'plot-connectivity-matrix', fig.height=8, fig.width=8, fig.cap="Figure 9. Connectivity matrix", out.width='80%'}
## Visualize connectivity matrix ----
gg_matrix(adour_matrix)
```


\

Optionally, we can use the function `matrix_to_edges_list()` to convert back the
connectivity matrix to edges list.

```{r 'connectivity-matrix-to-df'}
## Convert connectivity matrix to edges list ----
matrix_to_edges_list(adour_matrix)
```


\


## Nodes-by-edges matrix

The first object required by AEM is the **nodes-by-edges matrix**, i.e. a binary
matrix of dimensions `n x k`, where `n` is the number of nodes (in rows) and `k` the 
number of edges (in columns). This matrix represents the presence of a link 
influencing a site, directly or indirectly. It also gives the direction of the 
network.


\


The function `nodes_by_edges_matrix()` of the package `chessboard` is an alternative
to the function 
[`aem.build.binary()`](https://sdray.github.io/adespatial/reference/aem.build.binary.html) 
of the package [`adespatial`](https://cran.r-project.org/package=adespatial). 
This function creates the same output but works in a different way: it's only based on nodes 
labels (not on spatial coordinates). In addition, the function `nodes_by_edges_matrix()` 
adds labels to nodes and edges.



```{r 'nodes-by-edges-matrix'}
## Create nodes-by-edges matrix ----
adour_edges_mat <- nodes_by_edges_matrix(adour_edges)

## Content ----
names(adour_edges_mat)

## Header of the nodes-by-edges matrix ----
adour_edges_mat$"se.mat"[ , 1:15]

## Edges list ----
adour_edges_mat$"edges"
```

This function returns a list with the following elements:
  
  - `se.mat`: the nodes-by-edges matrix of dimensions `n x k`, where `n` is the 
number of nodes and `k` the number of edges (including the edge between the 
                                             fictitious origin and the first site);
- `edges`: a `data.frame` of edges list.


\


**Note:** a fictitious `origin` site (coded as `0`) has been created and represents
the first edge (coded as `E-01`) to the first site.


\


Let's plot our nodes-by-edges matrix.


```{r 'plot-nodes-by-edges-matrix', fig.height=4.5, fig.width=8, fig.cap="Figure 10. Nodes-by-edges matrix", out.width='80%'}
gg_matrix(adour_edges_mat$"se.mat")
```




\


## References

Blanchet F. G., Legendre P. & Borcard D. (2008) Modelling directional spatial 
processes in ecological data. **Ecological Modelling**, 215, 325-336. 
doi: [10.1016/j.ecolmodel.2008.04.001](https://doi.org/10.1016/j.ecolmodel.2008.04.001).

Dray S. _et al._ (2022) adespatial: Multivariate Multiscale Spatial Analysis. 
R package version 0.3-16, 
[https://CRAN.R-project.org/package=adespatial](https://CRAN.R-project.org/package=adespatial).
