---
title: "Chess movements"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Chess movements}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: inline
---

```{r include=FALSE}
knitr::opts_chunk$set(collapse  = TRUE,
                      comment   = "#>",
                      out.width = "50%",
                      dpi       = 96,
                      fig.align = "center")
```


```{r 'setup', echo=FALSE}
library("chessboard")
library("ggplot2")
library("patchwork")
```

```{r 'rerun-chunks', echo=FALSE}
eval_chunk <- TRUE
```



```{r 'fictitious-data', echo=FALSE}
sites <- expand.grid("transect" = 1:9, "quadrat" = 1:9)

nodes <- create_nodes_labels(data     = sites,
                             transect = "transect", 
                             quadrat  = "quadrat")

focus  <- "5-5"

transects_quadrats <- expand.grid("transect" = 1:3, "quadrat" = 1:5)
transects_only     <- data.frame("transect" = 1:3)
quadrats_only      <- data.frame("quadrat"  = 1:5)

nodes_transects_quadrats <- create_nodes_labels(data     = transects_quadrats,
                                                transect = "transect", 
                                                quadrat  = "quadrat")

nodes_transects_only <- create_nodes_labels(data     = transects_only,
                                            transect = "transect")

nodes_quadrats_only <- create_nodes_labels(data    = quadrats_only,
                                           quadrat = "quadrat")
```



The aim of the package `chessboard` is to make easier the creation of connectivity 
matrix, i.e. a binary matrix of dimensions `n x n`, where `n` is the number of 
**nodes** (sampling units) indicating the presence (`1`) or the absence (`0`) of
an **edge** (link) between pairs of nodes. Identifying edges between pairs of 
nodes requires the **detection of neighbors** according to neighborhood rules.


\


## Neighborhood rules


The package implements the following neighborhood rules:

- the **degree of neighborhood**: the number of adjacent nodes that will be used
to create direct edges. If `degree = 1`, only nodes directly adjacent (on a 
regular grid) to the focus node will be considered as neighbors.

- the **orientation of neighborhood**: can neighbors be detecting horizontally 
and/or vertically and/or diagonally? The package `chessboard` implements all possible
orientations derived from the chess game.

- the **direction of neighborhood**: does the sampling has a direction? This can
be particularly important in directed network (see below).


\


## Network properties



The package `chessboard` is designed to work with two-dimensional networks (i.e. 
sampling on a regular grid), where one dimension will be called **transect** 
and the other will be called **quadrat**. By convention, the dimension **transect** 
corresponds to the x-axis, and the **quadrat** corresponds to the y-axis (Fig. 1). 
The number of sampling units in both dimensions can be different.


\


```{r 'cb-transects-quadrats', eval=TRUE, fig.height=8, fig.width=5, echo=FALSE, fig.cap="Figure 1. Two-dimensional network"}
gg_chessboard(nodes_transects_quadrats)
```


\


It can also deal with one-dimensional sampling (either **transects-only** or 
**quadrats-only**). If the sampling has been conducted along one single 
dimension, a fictitious second dimension will be created (for the purpose of 
the package only). For instance, if the sampling has been conducted only 
through transects, a fictitious quadrat will be added with only one quadrat (Fig. 2).



```{r 'cb-transects-only', eval=eval_chunk, fig.height=2.1, fig.width=5, echo=FALSE, fig.cap="Figure 2. Transects-only network"}
gg_chessboard(nodes_transects_only)
```


\


On the other hand, if the sampling has been conducted only through quadrats, a
fictitious transect will be added with only one transect (Fig. 3).


```{r 'cb-quadrats-only', eval=eval_chunk, fig.height=8, fig.width=2.1, echo=FALSE, fig.cap="Figure 3. Quadrats-only network", out.width='20%'}
gg_chessboard(nodes_quadrats_only)
```


\


The network can be undirected or directed. If the network is **directed**, it will 
have (by default) these two orientations:

- from bottom to top for quadrats along transects, and
- from left to right for transects along quadrats.


It's important to note that, even the package `chessboard` handles spatial networks, 
it does not explicitly use spatial coordinates to find neighbors. Instead, it 
will identify neighbors only based on **nodes labels**. In other words, the order
of nodes labels will give the direction of the network.


\


The function `create_nodes_labels()` is used to create nodes labels. They 
will be of the form: `1-2`, where `1` is the identifier of 
the transect, and `2`, the identifier of the quadrat.


\


## Neighbors detection

Ignoring the spatial coordinates of the sampling design, we can consider it as
a **chessboard** of dimensions `t` x `q`, with `t`, the number of transects, 
and `q`, the number of quadrats.


\


### Degree of neighborhood

All functions for detecting neighbors (`pawn()`, `fool()`, etc.) have an 
argument `degree` that allow users to change the degree of neighborhood (Fig. 4).


\


```{r 'cb-degree', eval=eval_chunk, fig.height=4.3, fig.width=12, echo=FALSE, fig.cap="Figure 4. Degrees of neigborhood", out.width='100%'}
wizard_1 <- 
  gg_chessboard(nodes, "A. Degree of neighborhood of 1", "") + 
  geom_piece(nodes, focus) +
  geom_neighbors(nodes, wizard(nodes, focus, degree = 1))

wizard_2 <- 
  gg_chessboard(nodes, "B. Degree of neighborhood of 2", "") + 
  geom_piece(nodes, focus) +
  geom_neighbors(nodes, wizard(nodes, focus, degree = 2))

wizard_3 <- 
  gg_chessboard(nodes, "C. Degree of neighborhood of 3", "") + 
  geom_piece(nodes, focus) +
  geom_neighbors(nodes, wizard(nodes, focus, degree = 3))

(wizard_1 | wizard_2 | wizard_3)
```


\


### Movements


Different neighbors detection methods have been implemented in the package 
`chessboard`. All are derived from the chess game and allow user to define 
complex movements.


\



#### Pawn


The function `pawn()` can detect neighbors horizontally, i.e. among 
quadrats along a transect. User can change the default settings (Fig. 5A),
by adding directionality (`directed = TRUE`, Fig. 5B) and reversing the default
directionality (`directed = TRUE` and `reverse = TRUE`, Fig. 5C).


\


```{r 'cb-pawn', eval=eval_chunk, fig.height=4.3, fig.width=12, echo=FALSE, fig.cap="Figure 5. Pawn movements", out.width='100%'}
pawn_1 <- 
  gg_chessboard(nodes, "A. Undirected network", "") + 
  geom_piece(nodes, focus) +
  geom_neighbors(nodes, pawn(nodes, focus, degree = 4, directed = FALSE, reverse = FALSE))

pawn_2 <- 
  gg_chessboard(nodes, "B. Directed network", "") + 
  geom_piece(nodes, focus) +
  geom_neighbors(nodes, pawn(nodes, focus, degree = 4, directed = TRUE, reverse = FALSE))

pawn_3 <- 
  gg_chessboard(nodes, "C. Directed network (reverse)", "") + 
  geom_piece(nodes, focus) +
  geom_neighbors(nodes, pawn(nodes, focus, degree = 4, directed = TRUE, reverse = TRUE))

(pawn_1 | pawn_2 | pawn_3)
```


\


#### Fool

The function `fool()` can detect neighbors vertically, i.e. among 
transects along a quadrat. User can change the default settings (Fig. 6A),
by adding directionality (`directed = TRUE`, Fig. 6B) and reversing the default
directionality (`directed = TRUE` and `reverse = TRUE`, Fig. 6C).

\


```{r 'cb-fool', eval=eval_chunk, fig.height=4.3, fig.width=12, echo=FALSE, fig.cap="Figure 6. Fool movements", out.width='100%'}
fool_1 <- 
  gg_chessboard(nodes, "A. Undirected network", "") + 
  geom_piece(nodes, focus) +
  geom_neighbors(nodes, fool(nodes, focus, degree = 4, directed = FALSE, reverse = FALSE))

fool_2 <- 
  gg_chessboard(nodes, "B. Directed network", "") + 
  geom_piece(nodes, focus) +
  geom_neighbors(nodes, fool(nodes, focus, degree = 4, directed = TRUE, reverse = FALSE))

fool_3 <- 
  gg_chessboard(nodes, "C. Directed network (reverse)", "") + 
  geom_piece(nodes, focus) +
  geom_neighbors(nodes, fool(nodes, focus, degree = 4, directed = TRUE, reverse = TRUE))

(fool_1 | fool_2 | fool_3)
```


#### Rook


The function `rook()` can detect neighbors both horizontally and vertically, 
i.e. among transects along a quadrat and among quadrats along a transect. User 
can change the default settings (Fig. 7A), by adding directionality 
(`directed = TRUE`, Fig. 7B) and reversing the default directionality 
(`directed = TRUE` and `reverse = TRUE`, Fig. 7C).


\


```{r 'cb-rook', eval=eval_chunk, fig.height=4.3, fig.width=12, echo=FALSE, fig.cap="Figure 7. Rook movements", out.width='100%'}
rook_1 <- 
  gg_chessboard(nodes, "A. Undirected network", "") + 
  geom_piece(nodes, focus) +
  geom_neighbors(nodes, rook(nodes, focus, degree = 4, directed = FALSE, reverse = FALSE))

rook_2 <- 
  gg_chessboard(nodes, "B. Directed network", "") + 
  geom_piece(nodes, focus) +
  geom_neighbors(nodes, rook(nodes, focus, degree = 4, directed = TRUE, reverse = FALSE))

rook_3 <- 
  gg_chessboard(nodes, "C. Directed network (reverse)", "") + 
  geom_piece(nodes, focus) +
  geom_neighbors(nodes, rook(nodes, focus, degree = 4, directed = TRUE, reverse = TRUE))

(rook_1 | rook_2 | rook_3)
```


\


#### Bishop


The function `bishop()` can detect neighbors diagonally. User 
can change the default settings (Fig. 8A), by adding directionality 
(`directed = TRUE`, Fig. 8B) and reversing the default directionality 
(`directed = TRUE` and `reverse = TRUE`, Fig. 8C).

\


```{r 'cb-bishop', eval=eval_chunk, fig.height=4.3, fig.width=12, echo=FALSE, fig.cap="Figure 8. Bishop movements", out.width='100%'}
bishop_1 <- 
  gg_chessboard(nodes, "A. Undirected network", "") + 
  geom_piece(nodes, focus) +
  geom_neighbors(nodes, bishop(nodes, focus, degree = 4, directed = FALSE, reverse = FALSE))

bishop_2 <- 
  gg_chessboard(nodes, "B. Directed network", "") + 
  geom_piece(nodes, focus) +
  geom_neighbors(nodes, bishop(nodes, focus, degree = 4, directed = TRUE, reverse = FALSE))

bishop_3 <- 
  gg_chessboard(nodes, "C. Directed network (reverse)", "") + 
  geom_piece(nodes, focus) +
  geom_neighbors(nodes, bishop(nodes, focus, degree = 4, directed = TRUE, reverse = TRUE))

(bishop_1 | bishop_2 | bishop_3)
```


\


#### Bishop right

The function `bishop_right()` is a special case of the function `bishop()` and 
can detect neighbors along one diagonal (from bottom-left to top-right). User 
can change the default settings (Fig. 9A), by adding directionality 
(`directed = TRUE`, Fig. 9B) and reversing the default directionality 
(`directed = TRUE` and `reverse = TRUE`, Fig. 9C).

\


```{r 'cb-bishop-right', eval=eval_chunk, fig.height=4.3, fig.width=12, echo=FALSE, fig.cap="Figure 9. Bishop right movements", out.width='100%'}
bishop_right_1 <- 
  gg_chessboard(nodes, "A. Undirected network", "") + 
  geom_piece(nodes, focus) +
  geom_neighbors(nodes, bishop_right(nodes, focus, degree = 4, directed = FALSE, reverse = FALSE))

bishop_right_2 <- 
  gg_chessboard(nodes, "B. Directed network", "") + 
  geom_piece(nodes, focus) +
  geom_neighbors(nodes, bishop_right(nodes, focus, degree = 4, directed = TRUE, reverse = FALSE))

bishop_right_3 <- 
  gg_chessboard(nodes, "C. Directed network (reverse)", "") + 
  geom_piece(nodes, focus) +
  geom_neighbors(nodes, bishop_right(nodes, focus, degree = 4, directed = TRUE, reverse = TRUE))

(bishop_right_1 | bishop_right_2 | bishop_right_3)
```


\


#### Bishop left


The function `bishop_left()` is a special case of the function `bishop()` and 
can detect neighbors along one diagonal (from bottom-right to top-left). User 
can change the default settings (Fig. 10A), by adding directionality 
(`directed = TRUE`, Fig. 10B) and reversing the default directionality 
(`directed = TRUE` and `reverse = TRUE`, Fig. 10C).

\


```{r 'cb-bishop-left', eval=eval_chunk, fig.height=4.3, fig.width=12, echo=FALSE, fig.cap="Figure 10. Bishop left movements", out.width='100%'}
bishop_left_1 <- 
  gg_chessboard(nodes, "A. Undirected network", "") + 
  geom_piece(nodes, focus) +
  geom_neighbors(nodes, bishop_left(nodes, focus, degree = 4, directed = FALSE, reverse = FALSE))

bishop_left_2 <- 
  gg_chessboard(nodes, "B. Directed network", "") + 
  geom_piece(nodes, focus) +
  geom_neighbors(nodes, bishop_left(nodes, focus, degree = 4, directed = TRUE, reverse = FALSE))

bishop_left_3 <- 
  gg_chessboard(nodes, "C. Directed network (reverse)", "") + 
  geom_piece(nodes, focus) +
  geom_neighbors(nodes, bishop_left(nodes, focus, degree = 4, directed = TRUE, reverse = TRUE))

(bishop_left_1 | bishop_left_2 | bishop_left_3)
```


\


#### Knight

The function `knight()` has complex movement, derived from the chess game. User 
can change the default settings (Fig. 11A), by adding directionality 
(`directed = TRUE`, Fig. 11B) and reversing the default directionality 
(`directed = TRUE` and `reverse = TRUE`, Fig. 11C).

\


```{r 'cb-knight', eval=eval_chunk, fig.height=4.3, fig.width=12, echo=FALSE, fig.cap="Figure 11. Knight movements", out.width='100%'}
knight_1 <- 
  gg_chessboard(nodes, "A. Undirected network", "") + 
  geom_piece(nodes, focus) +
  geom_neighbors(nodes, knight(nodes, focus, degree = 4, directed = FALSE, reverse = FALSE))

knight_2 <- 
  gg_chessboard(nodes, "B. Directed network", "") + 
  geom_piece(nodes, focus) +
  geom_neighbors(nodes, knight(nodes, focus, degree = 4, directed = TRUE, reverse = FALSE))

knight_3 <- 
  gg_chessboard(nodes, "C. Directed network (reverse)", "") + 
  geom_piece(nodes, focus) +
  geom_neighbors(nodes, knight(nodes, focus, degree = 4, directed = TRUE, reverse = TRUE))

(knight_1 | knight_2 | knight_3)
```


\


#### Knight right

The function `knight_right()` is a special case of the function `knight()`. User 
can change the default settings (Fig. 12A), by adding directionality 
(`directed = TRUE`, Fig. 12B) and reversing the default directionality 
(`directed = TRUE` and `reverse = TRUE`, Fig. 12C).


\


```{r 'cb-knight-right', eval=eval_chunk, fig.height=4.3, fig.width=12, echo=FALSE, fig.cap="Figure 12. Knight right movements", out.width='100%'}
knight_right_1 <- 
  gg_chessboard(nodes, "A. Undirected network", "") + 
  geom_piece(nodes, focus) +
  geom_neighbors(nodes, knight_right(nodes, focus, degree = 4, directed = FALSE, reverse = FALSE))

knight_right_2 <- 
  gg_chessboard(nodes, "B. Directed network", "") + 
  geom_piece(nodes, focus) +
  geom_neighbors(nodes, knight_right(nodes, focus, degree = 4, directed = TRUE, reverse = FALSE))

knight_right_3 <- 
  gg_chessboard(nodes, "C. Directed network (reverse)", "") + 
  geom_piece(nodes, focus) +
  geom_neighbors(nodes, knight_right(nodes, focus, degree = 4, directed = TRUE, reverse = TRUE))

(knight_right_1 | knight_right_2 | knight_right_3)
```


\


#### Knight left

The function `knight_left()` is a special case of the function `knight()`. User 
can change the default settings (Fig. 13A), by adding directionality 
(`directed = TRUE`, Fig. 13B) and reversing the default directionality 
(`directed = TRUE` and `reverse = TRUE`, Fig. 13C).

\


```{r 'cb-knight-left', eval=eval_chunk, fig.height=4.3, fig.width=12, echo=FALSE, fig.cap="Figure 13. Knight left movements", out.width='100%'}
knight_left_1 <- 
  gg_chessboard(nodes, "A. Undirected network", "") + 
  geom_piece(nodes, focus) +
  geom_neighbors(nodes, knight_left(nodes, focus, degree = 4, directed = FALSE, reverse = FALSE))

knight_left_2 <- 
  gg_chessboard(nodes, "B. Directed network", "") + 
  geom_piece(nodes, focus) +
  geom_neighbors(nodes, knight_left(nodes, focus, degree = 4, directed = TRUE, reverse = FALSE))

knight_left_3 <- 
  gg_chessboard(nodes, "C. Directed network (reverse)", "") + 
  geom_piece(nodes, focus) +
  geom_neighbors(nodes, knight_left(nodes, focus, degree = 4, directed = TRUE, reverse = TRUE))

(knight_left_1 | knight_left_2 | knight_left_3)
```



\


#### Queen


The function `queen()` can detect neighbors horizontally, vertically and 
diagonally. It's a combination the `rook()` and the `bishop()`. User 
can change the default settings (Fig. 14A), by adding directionality 
(`directed = TRUE`, Fig. 14B) and reversing the default directionality 
(`directed = TRUE` and `reverse = TRUE`, Fig. 14C).

\


```{r 'cb-queen', eval=eval_chunk, fig.height=4.3, fig.width=12, echo=FALSE, fig.cap="Figure 14. Queen movements", out.width='100%'}
queen_1 <- 
  gg_chessboard(nodes, "A. Undirected network", "") + 
  geom_piece(nodes, focus) +
  geom_neighbors(nodes, queen(nodes, focus, degree = 4, directed = FALSE, reverse = FALSE))

queen_2 <- 
  gg_chessboard(nodes, "B. Directed network", "") + 
  geom_piece(nodes, focus) +
  geom_neighbors(nodes, queen(nodes, focus, degree = 4, directed = TRUE, reverse = FALSE))

queen_3 <- 
  gg_chessboard(nodes, "C. Directed network (reverse)", "") + 
  geom_piece(nodes, focus) +
  geom_neighbors(nodes, queen(nodes, focus, degree = 4, directed = TRUE, reverse = TRUE))

(queen_1 | queen_2 | queen_3)
```


\


#### Wizard

The function `wizard()` can detect neighbors in any direction. It's a 
combination the `queen()` and the `knight()`. User 
can change the default settings (Fig. 15A), by adding directionality 
(`directed = TRUE`, Fig. 15B) and reversing the default directionality 
(`directed = TRUE` and `reverse = TRUE`, Fig. 15C).
\


```{r 'cb-wizard', eval=eval_chunk, fig.height=4.3, fig.width=12, echo=FALSE, fig.cap="Figure 15. Wizard movements", out.width='100%'}
wizard_1 <- 
  gg_chessboard(nodes, "A. Undirected network", "") + 
  geom_piece(nodes, focus) +
  geom_neighbors(nodes, wizard(nodes, focus, degree = 4, directed = FALSE, reverse = FALSE))

wizard_2 <- 
  gg_chessboard(nodes, "B. Directed network", "") + 
  geom_piece(nodes, focus) +
  geom_neighbors(nodes, wizard(nodes, focus, degree = 4, directed = TRUE, reverse = FALSE))

wizard_3 <- 
  gg_chessboard(nodes, "C. Directed network (reverse)", "") + 
  geom_piece(nodes, focus) +
  geom_neighbors(nodes, wizard(nodes, focus, degree = 4, directed = TRUE, reverse = TRUE))

(wizard_1 | wizard_2 | wizard_3)
```


\


### Complex movements


It's possible to create complex movements by combining the previous movements
and by changing the arguments `degree`, `directed`, and `reverse`


```{r 'cb-custom', eval=eval_chunk, fig.height=4.3, fig.width=12, echo=FALSE, fig.cap="Figure 16. Custom movements", out.width='100%'}
wizard_1 <- 
  gg_chessboard(nodes, "Pawn & Bishop", "") + 
  geom_piece(nodes, focus) +
  geom_neighbors(nodes, pawn(nodes, focus, degree = 3, directed = TRUE, reverse = FALSE)) +
  geom_neighbors(nodes, bishop(nodes, focus, degree = 4, directed = TRUE, reverse = FALSE))

wizard_2 <- 
  gg_chessboard(nodes, "Knight L & Bishop L", "") + 
  geom_piece(nodes, focus) +
  geom_neighbors(nodes, knight_left(nodes, focus, degree = 4, directed = TRUE, reverse = FALSE)) +
  geom_neighbors(nodes, bishop_left(nodes, focus, degree = 4, directed = TRUE, reverse = FALSE))

wizard_3 <- 
  gg_chessboard(nodes, "Knight L, Bishop L, Pawn & Fool", "") + 
  geom_piece(nodes, focus) +
  geom_neighbors(nodes, knight_left(nodes, focus, degree = 4, directed = TRUE, reverse = FALSE)) +
  geom_neighbors(nodes, bishop_left(nodes, focus, degree = 4, directed = TRUE, reverse = FALSE)) +
  geom_neighbors(nodes, pawn(nodes, focus, degree = 4, directed = TRUE, reverse = FALSE)) +
  geom_neighbors(nodes, fool(nodes, focus, degree = 4, directed = TRUE, reverse = TRUE))

(wizard_1 | wizard_2 | wizard_3)
```
